---
title: "create input data"
author: "Luise A. Seeker"
date: "19/07/2020"
output: html_document
---
### Load libraries

```{r}

library(dplyr)
```

### read in data

```{r, echo=FALSE}
data<-read.csv("/Users/luise/Documents/Work/CattleTelomereLength/LongTelomereChangePaper/git/TelomereChangeInDairyCattle/CattleLongTelomereData.csv")

data$recoded_id <- as.factor(data$recoded_id)

names(data)

colnames(data)[19]<-"abs_diff_rltl_res"

names(data)
```



### Add sample year to data
```{r, echo = FALSE}
data$sample_year<-format(as.Date(data$sample_date, format="%d/%m/%Y"),"%Y")

names(data)
```

## Add lifelong RLTL dynamics measurements

### Add average residual RLTL 

```{r, echo = FALSE}

ET <- data.frame(recoded_id=sort(unique(data$recoded_id)))

ET$average_resid<-tapply(data$residual_rltl_t, data$recoded_id, mean)
nrow(ET)

data<-merge(data, ET, by="recoded_id", all=TRUE)
#names(data)
nrow(data)
names(data)


```

### Add mean RLTL change

```{r}
# This only works for data that is not NULL
redData<-subset(data, data$residual_rltl_tminus1 != "NULL")

eT <- data.frame(recoded_id=sort(unique(redData$recoded_id)))

eT$rltl_change<-tapply(as.numeric(paste(redData$diff_rltl_res)), redData$recoded_id, mean)
nrow(eT)


data<-merge(data, eT, by="recoded_id", all=TRUE)
#names(data)
nrow(data)
names(data)

```


### Add mean absolute RLTL change

```{r}
# Add mean absolute RLTL change:


ETA <- data.frame(recoded_id = sort(unique(redData$recoded_id)))

abs_diff<-as.numeric(paste(redData$abs_diff_rltl_res))

ETA$abs_rltl_change <- tapply(abs_diff, redData$recoded_id, mean)

nrow(ETA)


data<-merge(data, ETA, by="recoded_id", all=TRUE)
#names(data)
nrow(data)
names(data)

```



```{r}
#add column with factor levels for age ("young" for <= 1 year, "older" for > 1 year)

data$AgeGroup<-ifelse(data$age_d<365, "young", "older")

```

### Add RLTL at the age of 1

```{r}

rltl_1<-subset(data, data$age_y == 1)
nrow(rltl_1)

rltl_1_df<- data.frame(recoded_id = rltl_1$recoded_id, rltl_1_res = rltl_1$residual_rltl_t )

#some animals may have more than 1 measurement around 1 year of age... average for those

av_meas<-as.data.frame(tapply(rltl_1$residual_rltl_t , rltl_1$recoded_id, mean))
colnames(av_meas)<-"av_rltl_1_res"
av_meas$recoded_id <-rownames(av_meas)



data<-merge(data, av_meas, by = "recoded_id", all.x = TRUE)

nrow(data)
names(data)
```

### Calculate sample number per animal

```{r}
#calculate number of samples per animal
nuOfSamp<-data.frame(recoded_id=sort(data$recoded_id))

Sdf<- nuOfSamp %>% 
  group_by(recoded_id) %>%
  summarise(no_rows = length(recoded_id))


Sdf<-as.data.frame(Sdf)



names(Sdf)<-c("recoded_id", "SamplesPerAnimal")
Sdf$Animal_No<-c(1:305)

summary(Sdf$SamplesPerAnimal)

data<-merge(data, Sdf, by="recoded_id", all=TRUE)

nrow(data)
names(data)

```





```{r}


write.csv(data, "/Users/luise/Documents/Work/CattleTelomereLength/LongTelomereChangePaper/git/TelomereChangeInDairyCattle/LongTelomereDataCattle.csv" )


```


# Create calf input data

### Prepare data
```{r, echo = FALSE}

##############################
##############################
################################

# needs lots of work!!!!

df<-data.frame(recoded_id=unique(data$recoded_id))

RLTL0<-subset(data, data$age_y==0)
RLTL1<-subset(data, data$age_y==1)

df2<-data.frame(recoded_id=RLTL0$recoded_id, RLTL0=RLTL0$residual_rltl_t, S_date0=RLTL0$sample_date, herd_life= RLTL0$herd_life, cull_date=RLTL0$cull_date, dob=RLTL0$dob, tel_sample_no= RLTL0$tel_sample_no, sample_year0= RLTL0$sample_year)

df3<-data.frame(recoded_id=RLTL1$recoded_id, RLTL1=RLTL1$residual_rltl_t, S_date1=RLTL1$sample_date, sample_year= RLTL1$sample_year)

InData<-merge(df, df2, by="recoded_id")
InData2<-merge(InData, df3, by="recoded_id")

InData2$sample_interval= as.Date(InData2$S_date1, "%d/%m/%Y")-as.Date(InData2$S_date0, "%d/%m/%Y")
InData2$sInt<-as.numeric(InData2$sample_interval)
InData2$CutoffDate<-"01/01/2017"
InData2$Event<-ifelse(InData2$cull_date=="NULL",0,1)
InData2$TimeMeasure<-ifelse(InData2$Event==0, as.Date(InData2$CutoffDate, "%d/%m/%Y")- as.Date(InData2$dob, "%d/%m/%Y"), as.numeric(paste(InData2$herd_life)))
nrow(InData2)

InData2$ResDiff<- InData2$RLTL1-InData2$RLTL0
#InData2$ChangeRate<-InData2$ResDiff/InData2$sInt

#InData2$ResDiff_4G<-ifelse(InData2$ResDiff <= quantile(InData2$ResDiff)[2], 1, ifelse(InData2$ResDiff<= quantile(InData2$ResDiff)[3], 2, ifelse(InData2$ResDiff<= quantile(InData2$ResDiff)[4], 3, 4)))

InData2$ResDiff_3G<-ntile(InData2$ResDiff,3)

#InData2$ResDiff_2G<-ifelse(InData2$ResDiff<= median(InData2$ResDiff),1,2)


write.csv(InData2, "/Users/luise/Documents/Work/CattleTelomereLength/LongTelomereChangePaper/git/TelomereChangeInDairyCattle/first_year_long_tel_cattle.csv" )
```



### Generate Cox data
```{r}
redData<-subset(data, data$residual_rltl_tminus1 != "NULL")
uRedData<-redData[!duplicated(redData$recoded_id),] 



coxData<-data.frame(recoded_id=uRedData$recoded_id, dob=uRedData$dob, herd_life=uRedData$herd_life, cull_date=uRedData$cull_date, average_resid=uRedData$average_resid, rltl_change=uRedData$rltl_change, abs_rltl_change=uRedData$abs_rltl_change, amp_305_lactation=uRedData$amp_305_lactation, av_rltl_1_res=uRedData$av_rltl_1_res)


#extract maximum sample number per animal and create a data frame that only contains the recoded is and the associated maximum sample number.

sampleNo<-data.frame(recoded_id= redData$recoded_id, tel_sample_no= redData$tel_sample_no)


maxSN<-data.frame(tapply(sampleNo$tel_sample_no, sampleNo$recoded_id, function(x){which.max(x)}))
nrow(maxSN)


names(maxSN)<-"totalTELnumber"

rownames(maxSN)

maxSN$recoded_id<-rownames(maxSN)
rownames(maxSN)<-c()

head(maxSN)

```


```{r}

coxData<-merge(coxData, maxSN, by = "recoded_id")


nrow(coxData)


coxData$CutoffDate<-"01/01/2017"
coxData$Event<-ifelse(coxData$cull_date=="NULL", 0, 1)
coxData$TimeMeasure<-ifelse(coxData$Event==0, as.Date(coxData$CutoffDate, "%d/%m/%Y")- as.Date(coxData$dob, "%d/%m/%Y"), as.numeric(paste(coxData$herd_life)))

#add groups for all lifelong change measurements that are split at the median
#coxData$AverageResi_2G<-ifelse(coxData$average_resid <= median(coxData$average_resid), 1, 2)
#coxData$rltl_change_2G<-ifelse(coxData$rltl_change <= median (coxData$rltl_change), 1, 2 )
#coxData$abs_rltl_change_2G<-ifelse(coxData$abs_rltl_change <= median (coxData$abs_rltl_change), 1, 2 )

#split at quartiles
#Q1<-quantile(coxData$average_resid)

#coxData$AverageResi_4G<-ifelse(coxData$average_resid <= quantile(coxData$average_resid)[2], 1, ifelse(coxData$average_resid <= quantile(coxData$average_resid)[3], 2, ifelse(coxData$average_resid <= quantile(coxData$average_resid)[4], 3, 4)))

#coxData$rltl_change_4G<-ifelse(coxData$rltl_change <= quantile(coxData$rltl_change)[2], 1, ifelse(coxData$rltl_change <= quantile(coxData$rltl_change)[3], 2, ifelse(coxData$rltl_change<= quantile(coxData$rltl_change)[4], 3, 4)))

#coxData$abs_rltl_change_4G<-ifelse(coxData$abs_rltl_change <= quantile(coxData$abs_rltl_change)[2], 1, ifelse(coxData$abs_rltl_change <= quantile(coxData$abs_rltl_change)[3], 2, ifelse(coxData$abs_rltl_change<= quantile(coxData$abs_rltl_change)[4], 3, 4)))


coxData$average_resid_3G<-ntile(coxData$average_resid,3)
coxData$rltl_change_3G<-ntile(coxData$rltl_change,3)
coxData$abs_rltl_change_3G<-ntile(coxData$abs_rltl_change,3)


write.csv(coxData, "/Users/luise/Documents/Work/CattleTelomereLength/LongTelomereChangePaper/git/TelomereChangeInDairyCattle/cox_data.csv")
```


# Create a dataframe that excludes early measurements taken before the age of 3 months


```{r}

LateDF<-subset(data, data$age_m>3) #dataset with animals with more than two samples

red_data_l<-subset(LateDF, LateDF$residual_rltl_tminus1 != "NULL")

```


###### Lifetime RLTL dynamics measures need to be re-calculated without the early measurements
```{r}
red_data_ldf<-data.frame(recoded_id = red_data_l$recoded_id, dob=red_data_l$dob, herd_life = red_data_l$herd_life, cull_date = red_data_l$cull_date, amp_305_lactation = red_data_l$amp_305_lactation, av_rltl_1_res = red_data_l$av_rltl_1_res, residual_rltl_t =red_data_l$residual_rltl_t, diff_rltl_res=red_data_l$diff_rltl_res, amp_305_lactation=red_data_l$amp_305_lactation)

dfL <- data.frame(recoded_id=sort(unique(red_data_ldf$recoded_id)))
dfL$average_resid<-tapply(red_data_ldf$residual_rltl_t, red_data_ldf$recoded_id, mean)
dfL$rltl_change<-tapply(as.numeric(paste(red_data_ldf$diff_rltl_res)), red_data_ldf$recoded_id, mean)
dfL$abs_rltl_change<-tapply(abs(as.numeric(paste(red_data_ldf$diff_rltl_res))), red_data_ldf$recoded_id, mean)


red_data_l_new<-merge(red_data_ldf, dfL, by = "recoded_id")
# This dataframe contains longitudinal measurements excluding 

names(red_data_l_new)
nrow(red_data_l_new)

```


###### Calculate sample number per animal
```{r}

#calculate number of samples per animal
samp_no_late<-data.frame(recoded_id=sort(red_data_l_new$recoded_id))

sdf<- samp_no_late %>% 
  group_by(recoded_id) %>%
  summarise(no_rows = length(recoded_id))


sdf<-as.data.frame(sdf)



names(sdf)<-c("recoded_id", "SamplesPerAnimal")


summary(sdf$SamplesPerAnimal)

sdf<-subset(sdf, sdf$SamplesPerAnimal > 1)

red_data_l_new<-merge(red_data_l_new, sdf, by="recoded_id", all=TRUE)

nrow(red_data_l_new)
names(red_data_l_new)


write.csv(red_data_l_new, "/Users/luise/Documents/Work/CattleTelomereLength/LongTelomereChangePaper/git/TelomereChangeInDairyCattle/excluded_less_3_month.csv")
```

#### generate a dataframe for animals with more than 2 samples
```{r}

More2<-subset(data, data$SamplesPerAnimal >= 3) #dataset with animals with more than two samples
redMore2<-subset(More2, More2$residual_rltl_tminus1 != "NULL")

drops <- c("average_resid","rltl_change", "abs_rltl_change")
redMore2<-redMore2[ , !(names(redMore2) %in% drops)]

dfM2 <- data.frame(recoded_id=sort(unique(redMore2$recoded_id)))
dfM2$average_resid_M2<-tapply(redMore2$residual_rltl_t, redMore2$recoded_id, mean)
dfM2$rltl_change_M2<-tapply(as.numeric(paste(redMore2$diff_rltl_res)), redMore2$recoded_id, mean)
dfM2$abs_rltl_change_M2<-tapply(abs(as.numeric(paste(redMore2$diff_rltl_res))), redMore2$recoded_id, mean)


dfM2_df<-merge(redMore2, dfM2, by = "recoded_id")
# This dataframe contains longitudinal measurements excluding 

names(dfM2_df)
nrow(dfM2_df)


write.csv(dfM2_df, "/Users/luise/Documents/Work/CattleTelomereLength/LongTelomereChangePaper/git/TelomereChangeInDairyCattle/more_2_sampl.csv" )

```


###### Calculate sample number per animal
```{r}

#calculate number of samples per animal
samp_no_late<-data.frame(recoded_id=sort(red_data_l_new$recoded_id))

sdf<- samp_no_late %>% 
  group_by(recoded_id) %>%
  summarise(no_rows = length(recoded_id))


sdf<-as.data.frame(sdf)



names(sdf)<-c("recoded_id", "SamplesPerAnimal")


summary(sdf$SamplesPerAnimal)

sdf<-subset(sdf, sdf$SamplesPerAnimal > 1)

red_data_l_new<-merge(red_data_l_new, sdf, by="recoded_id", all=TRUE)

nrow(red_data_l_new)
names(red_data_l_new)
```


```{r}


deadD<-subset(data, data$herd_life != "NULL")
deadMore2<-subset(More2, More2$herd_life != "NULL")

redDeadD<-subset(redData, redData$herd_life != "NULL")

redMore2<-subset(More2, More2$residual_rltl_tminus1 != "NULL")
redDeadMore2<-subset(redMore2, redMore2$herd_life != "NULL")


uDeadData<-redDeadD[!duplicated(redDeadD$recoded_id),]


uDeadDataMore2<-redDeadMore2[!duplicated(redDeadMore2$recoded_id),]



```
